---
title: "ASD multi-omics tutorial: parallel and sequential Zentangler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASD multi-omics tutorial: parallel and sequential Zentangler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

This tutorial runs the two Zentangler mediation APIs on the ASD multi-omics
example data:

1. `fit_multiview_parallel_zentangler()` for pure parallel mediation.
2. `fit_sequential_zentangler()` for one explicit sequential route.

The ASD example uses a time contrast as the exposure:

```text
X = 0 at baseline T0
X = 1 at follow-up T1
```

and one symptom score as the outcome.

# Packages

```{r}
library(Zentangler)
library(MultiAssayExperiment)
library(SummarizedExperiment)
library(S4Vectors)
```

# Load The ASD Data

The tutorial uses the ASD files in `data/`. If you run this from another
working directory, set `ZENTANGLER_DATA_DIR` to the folder containing the ASD
`.RData` files.

```{r}
helper_file <- c(
  "asd-data-helpers.R",
  file.path("vignettes", "asd-data-helpers.R")
)
helper_file <- helper_file[file.exists(helper_file)][[1]]
source(helper_file)

asd <- load_asd_example_data()
names(asd$omics)
```

# Build One ASD MultiAssayExperiment

Here we build a `MultiAssayExperiment` for the `T1` contrast and the `ABC`
symptom outcome.

```{r}
mae_t1_abc <- build_asd_mae(
  omics = asd$omics,
  symptoms = asd$symptoms,
  contrast = "T1",
  symptom = "ABC"
)

mae_t1_abc
```

# 1. Pure Parallel Zentangler

The parallel API treats all selected mediators as candidate routes of the form:

```text
X -> mediator -> Y
```

There are no mediator-to-mediator links in this model.

```{r}
parallel_fit <- fit_multiview_parallel_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  method_preset = "fast_lasso",
  sis_n = 25,
  fusion_mode = "early",
  y_family = "gaussian",
  fdr_method = "BH",
  fdr_scope = "global",
  fill_nonfinite_zero = TRUE,
  seed = 1
)

zentangler_model_summary(parallel_fit)
zentangler_view_summary(parallel_fit, q_threshold = 0.25)
head(zentangler_top_mediators(parallel_fit, n = 10))
```

# 2. One Sequential Route

The sequential API runs one ordered route. For example:

```text
X -> species -> fecal_metabolites -> Y
```

The first layer is screened by exposure association. Adjacent layers are
screened by mediator-to-mediator correlation, and retained links are refit by
regression.

```{r}
sequential_fit <- fit_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  path_templates = list(
    species_to_fecal = c("species", "fecal_metabolites")
  ),
  sis_n = 25,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.5,
  fusion_mode = "late",
  lambda_choice = "lambda.min",
  y_family = "gaussian",
  seed = 1
)

sequential_paths <- zentangler_sequential_paths(sequential_fit)
sequential_fit$settings$route_set_type
head(sequential_paths, 20)
summarize_sequential_zentangler(sequential_fit)$model_summary
head(zentangler_sequential_edges(sequential_fit), 20)
head(zentangler_sequential_terminals(sequential_fit), 20)
```

# 3. Additional Sequential Route Scenarios

The ASD example has four modalities:

```{r}
names(asd$omics)
```

Because the sequential API fits one explicit ordered route at a time, different
biological hypotheses are run as separate fits. The helper below keeps the call
consistent across scenarios.

```{r}
run_asd_sequential_route <- function(route_name, route_views, sis_n = 25) {
  fit <- fit_sequential_zentangler(
    mae_t1_abc,
    x_var = "X",
    y_var = "Y",
    path_templates = stats::setNames(list(route_views), route_name),
    sis_n = sis_n,
    cor_method = "spearman",
    min_abs_cor = 0.15,
    cor_q_threshold = 0.5,
    fusion_mode = "late",
    lambda_choice = "lambda.min",
    glmnet_alpha = 0.5,
    y_family = "gaussian",
    seed = 1
  )

  paths <- zentangler_sequential_paths(fit)
  if (nrow(paths) > 0) {
    paths$route_name <- route_name
    paths$view_route <- paste(route_views, collapse = " -> ")
    paths <- paths[
      ,
      c("route_name", "view_route", setdiff(names(paths), c("route_name", "view_route"))),
      drop = FALSE
    ]
  }

  list(
    fit = fit,
    paths = paths,
    model_summary = summarize_sequential_zentangler(fit)$model_summary
  )
}
```

Here are several two-modality and three-modality routes using the same four ASD
views.

```{r}
asd_route_scenarios <- list(
  species_to_kos = c("species", "kos"),
  species_to_fecal = c("species", "fecal_metabolites"),
  species_to_plasma = c("species", "plasma_metabolites"),
  kos_to_fecal = c("kos", "fecal_metabolites"),
  kos_to_plasma = c("kos", "plasma_metabolites"),
  fecal_to_plasma = c("fecal_metabolites", "plasma_metabolites"),
  species_to_kos_to_fecal = c("species", "kos", "fecal_metabolites"),
  species_to_kos_to_plasma = c("species", "kos", "plasma_metabolites")
)

asd_route_fits <- Map(
  run_asd_sequential_route,
  names(asd_route_scenarios),
  asd_route_scenarios
)

route_model_summaries <- do.call(
  rbind,
  lapply(names(asd_route_fits), function(route_name) {
    out <- asd_route_fits[[route_name]]$model_summary
    out$route_name <- route_name
    out$view_route <- paste(asd_route_scenarios[[route_name]], collapse = " -> ")
    out[, c("route_name", "view_route", setdiff(names(out), c("route_name", "view_route"))), drop = FALSE]
  })
)

route_path_list <- Filter(
  function(x) is.data.frame(x) && nrow(x) > 0,
  lapply(asd_route_fits, `[[`, "paths")
)
route_paths <- if (length(route_path_list) > 0) do.call(rbind, route_path_list) else data.frame()

route_model_summaries
if (nrow(route_paths) > 0) {
  head(route_paths[order(route_paths$q_primary, route_paths$p_primary), ], 20)
} else {
  route_paths
}
```

You can also define broader sequential layers instead of one named view per
layer. This example treats `species` and `kos` as a microbiome layer and fecal
and plasma metabolites as a metabolite layer. The runnable example below uses
late fusion. If the retained terminal mediators span both metabolite views, you
can switch `fusion_mode` to `"intermediate"` to try the cooperative terminal
fusion mode.

```{r}
grouped_layer_fit <- fit_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  stage_views = list(
    microbiome = c("species", "kos"),
    metabolites = c("fecal_metabolites", "plasma_metabolites")
  ),
  sis_n = 12,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "late",
  lambda_choice = "lambda.min",
  glmnet_alpha = 0.5,
  coop_rho = 0.2,
  y_family = "gaussian",
  seed = 1
)

grouped_layer_paths <- zentangler_sequential_paths(grouped_layer_fit)
summarize_sequential_zentangler(grouped_layer_fit)$model_summary
if (nrow(grouped_layer_paths) > 0) {
  table(grouped_layer_paths$terminal_view)
  head(grouped_layer_paths[order(grouped_layer_paths$q_primary, grouped_layer_paths$p_primary), ], 20)
} else {
  grouped_layer_paths
}
```

# 4. Sequential Diagnostics and Bootstrap

The sequential API exposes the same kind of layers as the parallel API:
complete paths, retained mediator-to-mediator edges, terminal B-stage effects,
and run diagnostics.

```{r}
zentangler_sequential_diagnostics(sequential_fit)
```

For uncertainty on the full path score, use the fixed-path bootstrap. Increase
`bootstrap_repeats` for real analysis; the small value below is only for a quick
tutorial run.

```{r}
sequential_boot_fit <- fit_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  path_templates = list(
    species_to_fecal = c("species", "fecal_metabolites")
  ),
  sis_n = 12,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "early",
  lambda_choice = "lambda.min",
  b_inference = "debiased_lasso",
  path_inference = "bootstrap_score",
  bootstrap_repeats = 5,
  y_family = "gaussian",
  seed = 1
)

head(zentangler_sequential_paths(sequential_boot_fit), 20)
```

# Save Results

```{r}
out_dir <- file.path("tutorial_results", "asd_sequential_zentangler")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(
  zentangler_all_mediators(parallel_fit),
  file.path(out_dir, "parallel_all_mediators.csv"),
  row.names = FALSE
)

write.csv(
  sequential_paths,
  file.path(out_dir, "sequential_species_to_fecal_paths.csv"),
  row.names = FALSE
)

write.csv(
  route_model_summaries,
  file.path(out_dir, "sequential_route_scenario_model_summaries.csv"),
  row.names = FALSE
)

write.csv(
  route_paths,
  file.path(out_dir, "sequential_route_scenario_paths.csv"),
  row.names = FALSE
)

write.csv(
  grouped_layer_paths,
  file.path(out_dir, "sequential_grouped_microbiome_to_metabolite_paths.csv"),
  row.names = FALSE
)

saveRDS(
  list(
    parallel_fit = parallel_fit,
    sequential_fit = sequential_fit,
    asd_route_fits = asd_route_fits,
    grouped_layer_fit = grouped_layer_fit,
    sequential_boot_fit = sequential_boot_fit
  ),
  file.path(out_dir, "asd_sequential_zentangler_fits.rds")
)
```

# Interpreting The Output

For sequential route fits, the main path table includes:

| Column | Meaning |
|---|---|
| `route_set_type` | `sequential` |
| `mediator_path` | feature-level mediator path |
| `a` | exposure-to-first-mediator coefficient |
| `d_product` | product of mediator-to-mediator link coefficients |
| `b` | terminal mediator-to-outcome coefficient |
| `sequential_score` | `a * d_product * b` |
| `fusion_mode` | terminal fusion mode: `early`, `intermediate`, or `late` |
| `p_primary` | conservative path-level screening p-value |
| `q_primary` | global FDR-adjusted path-level value |

For the pure parallel model, use the usual Zentangler helpers:

```{r}
head(zentangler_all_mediators(parallel_fit))
zentangler_threshold_summary(parallel_fit, q_threshold = seq(0.05, 0.25, by = 0.05))
```
