---
title: "ASD multi-omics tutorial: HIMA2 and Zentangler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASD multi-omics tutorial: HIMA2 and Zentangler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
   warning = F, message = F
)
```

# Overview

This tutorial shows how to run two mediation workflows on the ASD multi-omics
example files in `data/`:

1. **HIMA2** as a single-view comparator, run independently for each modality.
2. **Zentangler** as a multiview mediation model over all available modalities.

The example files are:

```text
data/ASD_Species.RData
data/ASD_KOs.RData
data/ASD_fecalMetabolites.RData
data/ASD_plasmaMetabolites.RData
data/ASD_Symptoms.RData
```

# Packages

```{r}
library(Zentangler)
library(MultiAssayExperiment)
library(SummarizedExperiment)
library(S4Vectors)

# HIMA is needed only for the single-view HIMA2 comparator.
if (!requireNamespace("HIMA", quietly = TRUE)) {
  stop("Install HIMA before running the HIMA2 part of this tutorial.")
}
if (!exists("hima2", envir = asNamespace("HIMA"), inherits = FALSE)) {
  stop("The installed HIMA package does not expose HIMA::hima2().")
}
```

# Load The ASD Data

The omics `.RData` files each contain a `pcl` list with `features`, `metadata`,
and `features_metadata`. The symptom file contains a long table with columns
`ID`, `Symptom`, `Timepoint`, and `Score`.

```{r}
find_data_dir <- function() {
  env_dir <- Sys.getenv("ZENTANGLER_DATA_DIR", unset = "")
  candidates <- c(
    env_dir,
    "data",
    file.path("..", "data"),
    file.path(getwd(), "data"),
    file.path(dirname(getwd()), "data")
  )
  candidates <- candidates[nzchar(candidates)]
  hits <- candidates[dir.exists(candidates)]
  if (length(hits) == 0) {
    stop(
      "Could not find the ASD data directory. Set ZENTANGLER_DATA_DIR to ",
      "the folder containing ASD_Species.RData, ASD_KOs.RData, ",
      "ASD_fecalMetabolites.RData, ASD_plasmaMetabolites.RData, and ",
      "ASD_Symptoms.RData."
    )
  }
  normalizePath(hits[[1]], mustWork = TRUE)
}

data_dir <- find_data_dir()
data_dir

read_pcl <- function(path) {
  e <- new.env(parent = emptyenv())
  load(path, envir = e)
  if (!exists("pcl", envir = e, inherits = FALSE)) {
    stop("Expected object `pcl` in ", path)
  }
  get("pcl", envir = e)
}

omics_files <- c(
  species = "ASD_Species.RData",
  kos = "ASD_KOs.RData",
  fecal_metabolites = "ASD_fecalMetabolites.RData",
  plasma_metabolites = "ASD_plasmaMetabolites.RData"
)

omics <- lapply(file.path(data_dir, omics_files), read_pcl)
names(omics) <- names(omics_files)
symptoms <- read_pcl(file.path(data_dir, "ASD_Symptoms.RData"))$features
```

# Build A MultiAssayExperiment

For the ASD tutorial, the exposure is treatment time:

```text
X = 0 at baseline T0
X = 1 at follow-up T1 or T2
```

The outcome is one symptom score. The helper below builds a
`MultiAssayExperiment` for one contrast and one symptom. The returned MAE uses
subject-time IDs as primary sample IDs so the modalities and phenotype table are
aligned.

```{r}
normalize_omics_time <- function(x) {
  z <- toupper(trimws(as.character(x)))
  out <- rep(NA_character_, length(z))
  out[z %in% c("1", "0", "T0", "BASELINE")] <- "T0"
  out[z %in% c("3", "10", "T1", "10WK", "WEEK10", "MTT_10WK")] <- "T1"
  out[z %in% c("2YR", "2YR_FOLLOWUP", "T2", "YEAR2", "2Y")] <- "T2"
  out
}

normalize_symptom_time <- function(x) {
  z <- toupper(trimws(as.character(x)))
  out <- rep(NA_character_, length(z))
  out[z %in% c("T0", "BASELINE", "0")] <- "T0"
  out[z %in% c("T1", "10", "10WK", "WEEK10", "1")] <- "T1"
  out[z %in% c("T2", "2YR", "2YR_FOLLOWUP", "2")] <- "T2"
  out
}

make_primary_id <- function(subject_id, timepoint) {
  paste(as.character(subject_id), as.character(timepoint), sep = "__")
}

numeric_feature_matrix <- function(x, fill_zero = TRUE, scale_features = TRUE) {
  x <- as.data.frame(x, check.names = FALSE, stringsAsFactors = FALSE)
  mat <- as.matrix(data.frame(lapply(x, function(z) {
    suppressWarnings(as.numeric(as.character(z)))
  }), check.names = FALSE))
  colnames(mat) <- colnames(x)
  rownames(mat) <- rownames(x)
  storage.mode(mat) <- "numeric"
  if (fill_zero) mat[!is.finite(mat)] <- 0
  keep <- apply(mat, 2, function(z) stats::var(z, na.rm = TRUE) > 0)
  mat <- mat[, keep, drop = FALSE]
  if (scale_features && ncol(mat) > 0) {
    mat <- scale(mat)
    mat[!is.finite(mat)] <- 0
  }
  mat
}

build_asd_mae <- function(
  omics,
  symptoms,
  contrast = c("T1", "T2"),
  symptom = "abc",
  views = names(omics),
  fill_zero = TRUE,
  scale_features = TRUE
) {
  contrast <- match.arg(contrast)
  keep_times <- c("T0", contrast)

  sym <- as.data.frame(symptoms, check.names = FALSE, stringsAsFactors = FALSE)
  sym$time_norm <- normalize_symptom_time(sym$Timepoint)
  sym <- sym[sym$Symptom == symptom & sym$time_norm %in% keep_times, , drop = FALSE]
  sym$primary <- make_primary_id(sym$ID, sym$time_norm)
  sym$X <- as.integer(sym$time_norm == contrast)
  sym$Y <- suppressWarnings(as.numeric(as.character(sym$Score)))
  sym <- sym[is.finite(sym$Y), c("primary", "ID", "time_norm", "X", "Y"), drop = FALSE]
  rownames(sym) <- sym$primary

  experiments <- list()
  sample_map_rows <- list()

  for (view in views) {
    pcl <- omics[[view]]
    meta <- as.data.frame(pcl$metadata, check.names = FALSE, stringsAsFactors = FALSE)
    feat <- as.data.frame(pcl$features, check.names = FALSE, stringsAsFactors = FALSE)

    meta$time_norm <- normalize_omics_time(meta$Time_point)
    meta$primary <- make_primary_id(meta[["Subject ID"]], meta$time_norm)
    rownames(feat) <- meta$primary

    idx <- meta$time_norm %in% keep_times & meta$primary %in% sym$primary
    if (!any(idx)) next

    mat <- numeric_feature_matrix(
      feat[idx, , drop = FALSE],
      fill_zero = fill_zero,
      scale_features = scale_features
    )
    if (nrow(mat) < 4 || ncol(mat) < 2) next

    # SummarizedExperiment stores assays as features x samples.
    experiments[[view]] <- SummarizedExperiment::SummarizedExperiment(
      assays = list(abundance = t(mat))
    )
    sample_map_rows[[view]] <- data.frame(
      assay = view,
      primary = rownames(mat),
      colname = rownames(mat),
      stringsAsFactors = FALSE
    )
  }

  if (length(experiments) == 0) {
    stop("No usable omics views were available for contrast=", contrast,
         " and symptom=", symptom)
  }

  primary_ids <- Reduce(union, lapply(sample_map_rows, `[[`, "primary"))
  pheno <- sym[intersect(rownames(sym), primary_ids), , drop = FALSE]
  sample_map <- do.call(rbind, sample_map_rows)
  sample_map <- sample_map[sample_map$primary %in% rownames(pheno), , drop = FALSE]
  sample_map <- S4Vectors::DataFrame(
    assay = sample_map$assay,
    primary = sample_map$primary,
    colname = sample_map$colname
  )

  MultiAssayExperiment::MultiAssayExperiment(
    experiments = experiments,
    colData = S4Vectors::DataFrame(
      X = pheno$X,
      Y = pheno$Y,
      subject_id = pheno$ID,
      timepoint = pheno$time_norm,
      row.names = rownames(pheno)
    ),
    sampleMap = sample_map
  )
}

mae_t1_abc <- build_asd_mae(
  omics = omics,
  symptoms = symptoms,
  contrast = "T1",
  symptom = "ABC"
)
mae_t1_abc
```

# Run Zentangler

Zentangler fits all selected views in one multiview model. For quick iteration,
start with the fast lasso preset and a modest SIS screen. Increase `sis_n`, use
`method_preset = "full_exploratory"`, or add bootstrap repeats for a larger
analysis.

The ASD example is a time-point contrast, so the call below uses
`study_design = "time"` instead of passing `x_var = "X"` directly. Zentangler
encodes `timepoint == "T0"` as 0 and `timepoint == "T1"` as 1 internally. The
explicit `X` column is still kept in the MAE because the HIMA2 comparator helper
below uses it directly.

```{r}
zt_fit <- fit_multiview_parallel_zentangler(
  mae_t1_abc,
  y_var = "Y",
  study_design = "time",
  time_var = "timepoint",
  time_ref = "T0",
  time_compare = "T1",
  covariates = NULL,
  glmnet_alpha = 0.5,
  sis_n = 50,
  sis_rank = "abs_a",
  screen_method = "sis",
  fusion_mode = "late",
  y_family = "gaussian",
  fdr_method = "BH",
  fdr_scope = "global",
  fill_nonfinite_zero = TRUE,
  seed = 1
)

zentangler_model_summary(zt_fit)
zentangler_view_summary(zt_fit, q_threshold = 0.25)
zentangler_top_mediators(zt_fit, n = 20)
zentangler_active_mediators(zt_fit, q_threshold = 0.25)
```

Save the main outputs:

```{r}
out_dir <- file.path("tutorial_results", "zentangler_t1_abc")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(zentangler_all_mediators(zt_fit),
          file.path(out_dir, "zentangler_all_mediators.csv"),
          row.names = FALSE)
write.csv(zentangler_active_mediators(zt_fit, q_threshold = 0.25),
          file.path(out_dir, "zentangler_active_mediators_q0p25.csv"),
          row.names = FALSE)
write.csv(zentangler_view_summary(zt_fit, q_threshold = 0.25),
          file.path(out_dir, "zentangler_view_summary_q0p25.csv"),
          row.names = FALSE)
saveRDS(zt_fit, file.path(out_dir, "zentangler_fit.rds"))
```

# Run HIMA2 One View At A Time

HIMA2 is a single-view method, so the comparator loops over the experiments in
the MAE and runs one model per modality. The helper below adapts to the two main
HIMA2 calling styles seen across HIMA versions.

```{r}
mae_block <- function(mae, view) {
  extract_block <- getFromNamespace("zentangler_mae_extract_block", "Zentangler")
  extract_block(
    mae,
    experiment_name = view,
    fill_nonfinite_zero = TRUE
  )
}

run_hima2_one_view <- function(M, pheno, penalty = "DBlasso") {
  fun <- get("hima2", envir = asNamespace("HIMA"))
  argn <- names(formals(fun))
  data_M <- as.data.frame(M, check.names = FALSE)

  if ("formula" %in% argn) {
    call_args <- list(formula = Y ~ X)
    if ("data.pheno" %in% argn) call_args$data.pheno <- pheno
    if ("data.M" %in% argn) call_args$data.M <- data_M
    if ("mediator.type" %in% argn) call_args$mediator.type <- "gaussian"
    if ("mediator.family" %in% argn) call_args$mediator.family <- "gaussian"
    if ("outcome.family" %in% argn) call_args$outcome.family <- "gaussian"
    if ("penalty" %in% argn) call_args$penalty <- penalty
    if ("scale" %in% argn) call_args$scale <- FALSE
    if ("verbose" %in% argn) call_args$verbose <- FALSE
    if ("topN" %in% argn) call_args$topN <- ncol(data_M)
    if ("sigcut" %in% argn) call_args$sigcut <- 1
    if ("Sigcut" %in% argn) call_args$Sigcut <- 1
    return(do.call(fun, call_args))
  }

  call_args <- list(X = pheno$X, Y = pheno$Y, M = as.matrix(M))
  if ("penalty" %in% argn) call_args$penalty <- penalty
  if ("topN" %in% argn) call_args$topN <- ncol(M)
  if ("sigcut" %in% argn) call_args$sigcut <- 1
  if ("Sigcut" %in% argn) call_args$Sigcut <- 1
  do.call(fun, call_args)
}

standardize_hima2_table <- function(fit, view, rep = NA_integer_) {
  tab <- tryCatch({
    if (inherits(fit, "hima") && is.list(fit)) {
      lens <- vapply(fit, length, integer(1))
      keep <- names(fit)[lens == max(lens) & lens > 0]
      as.data.frame(fit[keep], check.names = FALSE, stringsAsFactors = FALSE)
    } else {
      as.data.frame(fit, check.names = FALSE)
    }
  }, error = function(e) data.frame())
  if (nrow(tab) == 0) return(tab)

  if ("ID" %in% names(tab)) {
    tab$mediator <- as.character(tab$ID)
  } else {
    tab$mediator <- rownames(tab)
  }
  rownames(tab) <- NULL
  names(tab) <- make.names(names(tab), unique = TRUE)

  p_col <- intersect(c("p.value", "p.value.", "p", "P.value", "raw.p", "raw_p"), names(tab))
  q_col <- intersect(c("BH.FDR", "BH_FDR", "q.value", "qvalue", "FDR"), names(tab))
  a_col <- intersect(c("alpha", "a", "alpha_hat"), names(tab))
  b_col <- intersect(c("beta", "b", "beta_hat"), names(tab))
  ab_col <- intersect(c("alpha.beta", "alpha.beta.", "alpha.beta.."), names(tab))

  tab$p_primary <- if (length(p_col) > 0) as.numeric(tab[[p_col[[1]]]]) else NA_real_
  tab$q_hima_reported <- if (length(q_col) > 0) as.numeric(tab[[q_col[[1]]]]) else NA_real_
  tab$a <- if (length(a_col) > 0) as.numeric(tab[[a_col[[1]]]]) else NA_real_
  tab$b <- if (length(b_col) > 0) as.numeric(tab[[b_col[[1]]]]) else NA_real_
  tab$score <- if (length(ab_col) > 0) as.numeric(tab[[ab_col[[1]]]]) else tab$a * tab$b
  tab$abs_score <- abs(tab$score)
  tab$rep <- rep
  tab$method <- "hima2"
  tab$view <- view
  tab$omics <- view
  tab
}

run_hima2_mae <- function(mae,
                          x_var = "X",
                          y_var = "Y",
                          rep = NA_integer_,
                          penalty = "DBlasso") {
  pheno <- as.data.frame(SummarizedExperiment::colData(mae))
  if (!(x_var %in% colnames(pheno))) stop("x_var not found in MAE colData: ", x_var)
  if (!(y_var %in% colnames(pheno))) stop("y_var not found in MAE colData: ", y_var)
  pheno$X <- suppressWarnings(as.numeric(as.character(pheno[[x_var]])))
  pheno$Y <- suppressWarnings(as.numeric(as.character(pheno[[y_var]])))
  pheno <- pheno[is.finite(pheno$X) & is.finite(pheno$Y), , drop = FALSE]

  results <- list()
  for (view in names(MultiAssayExperiment::experiments(mae))) {
    M <- mae_block(mae, view)
    common <- intersect(rownames(pheno), rownames(M))
    M <- M[common, , drop = FALSE]
    pheno_view <- pheno[common, , drop = FALSE]
    if (length(unique(pheno_view$X)) < 2 || nrow(pheno_view) < 8 || ncol(M) < 2) {
      next
    }

    fit <- run_hima2_one_view(M, pheno_view, penalty = penalty)
    results[[view]] <- standardize_hima2_table(fit, view, rep = rep)
  }

  out <- if (length(results) > 0) do.call(rbind, results) else data.frame()
  if (nrow(out) > 0 && "p_primary" %in% names(out)) {
    out$p_primary <- as.numeric(unlist(out$p_primary, use.names = FALSE))
    out$q_primary_global <- p.adjust(out$p_primary, method = "BH")
    out$q_primary <- out$q_primary_global
    out$q_primary_within_view <- ave(
      out$p_primary,
      out$view,
      FUN = function(p) p.adjust(p, method = "BH")
    )
  }
  rownames(out) <- NULL
  out
}

hima2_tab <- run_hima2_mae(mae_t1_abc)
hima2_sig <- hima2_tab[
  is.finite(as.numeric(hima2_tab$q_primary_global)) &
    as.numeric(hima2_tab$q_primary_global) < 0.25,
  ,
  drop = FALSE
]

hima2_sig <- hima2_sig[
  order(as.numeric(hima2_sig$q_primary_global)),
  ,
  drop = FALSE
]

head(hima2_sig, 20)
```

Save HIMA2 outputs:

```{r}
out_dir <- file.path("tutorial_results", "hima2_t1_abc")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(hima2_tab,
          file.path(out_dir, "hima2_all_mediators.csv"),
          row.names = FALSE)
write.csv(subset(hima2_tab, q_primary_global <= 0.25),
          file.path(out_dir, "hima2_active_mediators_q0p25.csv"),
          row.names = FALSE)
```

# Run Multiple Symptoms And Contrasts

The same helper can run over symptoms and treatment contrasts. Fecal and plasma
metabolite views have baseline and T1 samples. Species and KOs also include T2
samples in these example objects.

```{r}
symptom_names <- sort(unique(symptoms$Symptom))
contrasts <- c("T1", "T2")

all_zentangler <- list()
all_hima2 <- list()
all_errors <- character(0)

for (contrast in contrasts) {
  for (symptom in symptom_names) {
    key <- paste(contrast, symptom, sep = "__")

    mae <- tryCatch(
      build_asd_mae(omics, symptoms, contrast = contrast, symptom = symptom),
      error = function(e) {
        all_errors <<- c(all_errors, paste0(key, ": MAE build failed: ", conditionMessage(e)))
        NULL
      }
    )
    if (is.null(mae)) next

    pheno <- as.data.frame(SummarizedExperiment::colData(mae))
    if (length(unique(pheno$X[is.finite(pheno$X)])) < 2) {
      all_errors <- c(all_errors, paste0(key, ": skipped because X has fewer than two levels."))
      next
    }

    zt <- tryCatch(
      fit_multiview_parallel_zentangler(
        mae,
        y_var = "Y",
        study_design = "time",
        time_var = "timepoint",
        time_ref = "T0",
        time_compare = contrast,
        covariates = NULL,
        glmnet_alpha = 0.5,
        sis_n = 50,
        sis_rank = "abs_a",
        screen_method = "sis",
        fusion_mode = "late",
        y_family = "gaussian",
        fdr_method = "BH",
        fdr_scope = "global",
        effect_x0 = 0,
        effect_x1 = 1,
        fill_nonfinite_zero = TRUE,
        seed = 1
      ),
      error = function(e) {
        all_errors <<- c(all_errors, paste0(key, ": Zentangler failed: ", conditionMessage(e)))
        NULL
      }
    )

    if (!is.null(zt)) {
      zt_tab <- zentangler_all_mediators(zt)
      zt_tab$contrast <- contrast
      zt_tab$symptom <- symptom
      all_zentangler[[key]] <- zt_tab
    }

    h2_tab <- tryCatch(run_hima2_mae(mae), error = function(e) data.frame())
    if (nrow(h2_tab) > 0) {
      h2_tab$contrast <- contrast
      h2_tab$symptom <- symptom
      all_hima2[[key]] <- h2_tab
    }
  }
}

zentangler_all <- if (length(all_zentangler)) do.call(rbind, all_zentangler) else data.frame()
hima2_all <- if (length(all_hima2)) do.call(rbind, all_hima2) else data.frame()

all_errors
```

# Zentangler Simulations

Zentangler also includes SIMMBA/InterSIM-style simulation helpers. This is useful
for checking that a method recovers known mediators before running the real ASD
analysis.

```{r, fig.width=8, fig.align='center', fig.height=6}
set.seed(1)

sim <- gen_simmba(
  nsample = 500,
  nrep =5,
  snr = 1,
  n.pathways = 30,
  TIE = rep(1, 3),
  outcome.type = "continuous",
  ygen.mode = "LM",
  seed = 1
)

zentangler_sim_recovery <- list()
hima2_sim_recovery <- list()
zentangler_sim_summaries <- list()
hima2_sim_results <- list()

for (rep_i in seq_along(sim$trainMae)) {
  sim_mae <- sim$trainMae[[rep_i]]
  truth <- sim$truthDat[[rep_i]]
  truth_key <- zentangler_truth_key(truth)

  sim_fit <- fit_multiview_parallel_zentangler(
    sim_mae,
    x_var = "A",
    y_var = "Y",
    covariates = NULL,
    glmnet_alpha = 0.5,
    sis_n = 50,
    sis_rank = "abs_a",
    screen_method = "sis",
    fusion_mode = "late",
    y_family = "gaussian",
    fdr_method = "BH",
    fdr_scope = "global",
    effect_x0 = 0,
    effect_x1 = 1,
    fill_nonfinite_zero = TRUE,
    seed = rep_i
  )

  zentangler_tab <- zentangler_all_mediators(sim_fit)
  zentangler_sim_summaries[[rep_i]] <- summarize_zentangler(
    sim_fit,
    q_threshold = seq(0.05, 0.25, by = 0.05)
  )
  zentangler_sim_recovery[[rep_i]] <- zentangler_score_truth_recovery(
    zentangler_tab,
    truth_key = truth_key,
    rep = rep_i,
    fusion_mode = "late",
    q_threshold = 0.25,
    q_col = "q_primary",
    fdr_scope = "global",
    top_n = 50
  )

  hima2_tab <- run_hima2_mae(
    sim_mae,
    x_var = "A",
    y_var = "Y",
    rep = rep_i
  )
  hima2_sim_results[[rep_i]] <- hima2_tab
  hima2_sim_recovery[[rep_i]] <- zentangler_score_truth_recovery(
    hima2_tab,
    truth_key = truth_key,
    rep = rep_i,
    fusion_mode = "singleview_hima2",
    q_threshold = 0.25,
    q_col = "q_primary",
    fdr_scope = "global",
    top_n = 50
  )
}

zentangler_recovery <- do.call(rbind, zentangler_sim_recovery)
hima2_recovery <- do.call(rbind, hima2_sim_recovery)

zentangler_recovery
hima2_recovery

sim_recovery_plot <- rbind(
  transform(zentangler_recovery, method = "Zentangler"),
  transform(hima2_recovery, method = "HIMA2")
)

plot_metric_by_method <- function(df, metric, ylab = metric) {
  boxplot(
    stats::as.formula(paste(metric, "~ method")),
    data = df,
    ylab = ylab,
    xlab = "",
    col = c("#4C78A8", "#F58518"),
    border = "#333333"
  )
  stripchart(
    stats::as.formula(paste(metric, "~ method")),
    data = df,
    vertical = TRUE,
    method = "jitter",
    pch = 16,
    col = "#00000066",
    add = TRUE
  )
}

op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot_metric_by_method(sim_recovery_plot, "recall", "Recall at q <= 0.25")
plot_metric_by_method(sim_recovery_plot, "precision", "Precision at q <= 0.25")
plot_metric_by_method(sim_recovery_plot, "fdr", "Observed FDR at q <= 0.25")
plot_metric_by_method(sim_recovery_plot, "top50_precision", "Top-50 precision")
par(op)

plot(
  zentangler_recovery$rep,
  zentangler_recovery$n_active,
  type = "b",
  pch = 16,
  col = "#4C78A8",
  ylim = range(c(zentangler_recovery$n_active, hima2_recovery$n_active), na.rm = TRUE),
  xlab = "Simulation replicate",
  ylab = "Active mediators at q <= 0.25"
)
lines(hima2_recovery$rep, hima2_recovery$n_active, type = "b", pch = 16, col = "#F58518")
legend(
  "topright",
  legend = c("Zentangler", "HIMA2"),
  col = c("#4C78A8", "#F58518"),
  pch = 16,
  lty = 1,
  bty = "n"
)

```
