---
title: "ASD multi-omics tutorial: parallel, sequential, and mixed Zentangler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASD multi-omics tutorial: parallel, sequential, and mixed Zentangler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

This tutorial runs the three Zentangler mediation APIs on the ASD multi-omics
example data:

1. `fit_multiview_parallel_zentangler()` for pure parallel mediation.
2. `fit_sequential_zentangler()` for one explicit sequential route.
3. `fit_parallel_sequential_zentangler()` for a user-specified set of routes
   that may contain both sequential and one-view parallel routes.

The ASD example uses a time contrast as the exposure:

```text
X = 0 at baseline T0
X = 1 at follow-up T1
```

and one symptom score as the outcome.

# Packages

```{r}
library(Zentangler)
library(MultiAssayExperiment)
library(SummarizedExperiment)
library(S4Vectors)
```

# Load The ASD Data

The tutorial uses the ASD files in `data/`. If you run this from another
working directory, set `ZENTANGLER_DATA_DIR` to the folder containing the ASD
`.RData` files.

```{r}
helper_file <- c(
  "asd-data-helpers.R",
  file.path("vignettes", "asd-data-helpers.R")
)
helper_file <- helper_file[file.exists(helper_file)][[1]]
source(helper_file)

asd <- load_asd_example_data()
names(asd$omics)
```

# Build One ASD MultiAssayExperiment

Here we build a `MultiAssayExperiment` for the `T1` contrast and the `ABC`
symptom outcome.

```{r}
mae_t1_abc <- build_asd_mae(
  omics = asd$omics,
  symptoms = asd$symptoms,
  contrast = "T1",
  symptom = "ABC"
)

mae_t1_abc
```

# 1. Pure Parallel Zentangler

The parallel API treats all selected mediators as candidate routes of the form:

```text
X -> mediator -> Y
```

There are no mediator-to-mediator links in this model.

```{r}
parallel_fit <- fit_multiview_parallel_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  method_preset = "fast_lasso",
  sis_n = 25,
  fusion_mode = "early",
  y_family = "gaussian",
  fdr_method = "BH",
  fdr_scope = "global",
  fill_nonfinite_zero = TRUE,
  seed = 1
)

zentangler_model_summary(parallel_fit)
zentangler_view_summary(parallel_fit, q_threshold = 0.25)
head(zentangler_top_mediators(parallel_fit, n = 10))
```

# 2. One Sequential Route

The sequential API runs one ordered route. For example:

```text
X -> species -> fecal_metabolites -> Y
```

The first layer is screened by exposure association. Adjacent layers are
screened by mediator-to-mediator correlation, and retained links are refit by
regression.

```{r}
sequential_fit <- fit_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  path_templates = list(
    species_to_fecal = c("species", "fecal_metabolites")
  ),
  sis_n = 40,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "late",
  lambda_choice = "lambda.min",
  y_family = "gaussian",
  seed = 1
)

sequential_paths <- zentangler_sequential_paths(sequential_fit)
sequential_fit$settings$route_set_type
head(sequential_paths, 20)
summarize_sequential_zentangler(sequential_fit)$model_summary
head(zentangler_sequential_edges(sequential_fit), 20)
head(zentangler_sequential_terminals(sequential_fit), 20)
```

# 3. Parallel Plus Sequential Route Set

The mixed API lets you specify exactly which biological routes to test. It does
not enumerate all possible combinations.

The example below runs three routes in parallel:

```text
X -> species -> fecal_metabolites -> Y
X -> kos -> plasma_metabolites -> Y
X -> plasma_metabolites -> Y
```

The first two are sequential routes. The third is a one-view parallel route
included in the same route set.

```{r}
mixed_fit <- fit_parallel_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  path_templates = list(
    species_to_fecal = c("species", "fecal_metabolites"),
    kos_to_plasma = c("kos", "plasma_metabolites"),
    plasma_direct = c("plasma_metabolites")
  ),
  sis_n = 40,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "early",
  lambda_choice = "lambda.min",
  y_family = "gaussian",
  seed = 1
)

mixed_paths <- zentangler_sequential_paths(mixed_fit)
mixed_fit$settings$route_set_type
table(mixed_paths$route_type)
head(mixed_paths, 20)
summarize_sequential_zentangler(mixed_fit)$threshold_summary
```

The sequential APIs now expose the same kind of layers as the parallel API:
complete paths, retained mediator-to-mediator edges, terminal B-stage effects,
and run diagnostics.

```{r}
zentangler_sequential_diagnostics(mixed_fit)
```

For uncertainty on the full path score, use the fixed-path bootstrap. Increase
`bootstrap_repeats` for real analysis; the small value below is only for a quick
tutorial run.

```{r}
mixed_boot_fit <- fit_parallel_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  path_templates = list(
    species_to_fecal = c("species", "fecal_metabolites"),
    plasma_direct = c("plasma_metabolites")
  ),
  sis_n = 12,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "early",
  lambda_choice = "lambda.min",
  b_inference = "debiased_lasso",
  path_inference = "bootstrap_score",
  bootstrap_repeats = 5,
  y_family = "gaussian",
  seed = 1
)

head(zentangler_sequential_paths(mixed_boot_fit), 20)
```

# Binary and Survival Sequential Outcomes

The same route APIs can be used with binary and survival outcomes. The ASD
example data here use continuous symptom scores, so the following chunks create
small derived outcomes only to demonstrate the calls.

```{r}
mae_binary_demo <- mae_t1_abc
colData(mae_binary_demo)$Y_binary <- as.numeric(
  colData(mae_binary_demo)$Y > stats::median(colData(mae_binary_demo)$Y, na.rm = TRUE)
)

binary_seq_fit <- fit_sequential_zentangler(
  mae_binary_demo,
  x_var = "X",
  y_var = "Y_binary",
  path_templates = list(species_to_fecal = c("species", "fecal_metabolites")),
  sis_n = 20,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "early",
  lambda_choice = "lambda.min",
  y_family = "binomial",
  b_inference = "debiased_logistic_lasso",
  seed = 1
)

head(zentangler_sequential_paths(binary_seq_fit), 20)
unique(zentangler_sequential_terminals(binary_seq_fit)$b_inference_method)
```

Survival outcomes use a time column and a 0/1 event column. Intermediate fusion
is currently Gaussian-only, so use early or late fusion for survival route fits.

```{r}
mae_survival_demo <- mae_t1_abc
set.seed(1)
colData(mae_survival_demo)$time <- seq_len(ncol(mae_survival_demo)) + stats::runif(ncol(mae_survival_demo))
colData(mae_survival_demo)$status <- as.numeric(
  colData(mae_survival_demo)$Y > stats::median(colData(mae_survival_demo)$Y, na.rm = TRUE)
)

survival_seq_fit <- fit_sequential_zentangler(
  mae_survival_demo,
  x_var = "X",
  y_var = NULL,
  path_templates = list(species_to_fecal = c("species", "fecal_metabolites")),
  sis_n = 20,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "early",
  lambda_choice = "lambda.min",
  y_family = "survival",
  survival_time_var = "time",
  survival_event_var = "status",
  b_inference = "debiased_cox_lasso",
  seed = 1
)

head(zentangler_sequential_paths(survival_seq_fit), 20)
unique(zentangler_sequential_terminals(survival_seq_fit)$b_inference_method)
```

# Mixed Routes With Intermediate Fusion

Use `fusion_mode = "intermediate"` when the terminal mediator-to-outcome stage
should use Zentangler's cooperative/intermediate learner. This requires at least
two terminal views across the requested route set. In the example below, the
terminal views are `fecal_metabolites` and `plasma_metabolites`.

```{r}
mixed_intermediate_fit <- fit_parallel_sequential_zentangler(
  mae_t1_abc,
  x_var = "X",
  y_var = "Y",
  path_templates = list(
    species_to_fecal = c("species", "fecal_metabolites"),
    kos_to_plasma = c("kos", "plasma_metabolites")
  ),
  sis_n = 8,
  cor_method = "spearman",
  min_abs_cor = 0.15,
  cor_q_threshold = 0.75,
  fusion_mode = "intermediate",
  lambda_choice = "lambda.min",
  coop_rho = 0.2,
  y_family = "gaussian",
  seed = 1
)

mixed_intermediate_paths <- zentangler_sequential_paths(mixed_intermediate_fit)
mixed_intermediate_fit$settings$route_set_type
table(mixed_intermediate_paths$terminal_view)
head(mixed_intermediate_paths, 20)
```

# Save Results

```{r}
out_dir <- file.path("tutorial_results", "asd_sequential_zentangler")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(
  zentangler_all_mediators(parallel_fit),
  file.path(out_dir, "parallel_all_mediators.csv"),
  row.names = FALSE
)

write.csv(
  sequential_paths,
  file.path(out_dir, "sequential_species_to_fecal_paths.csv"),
  row.names = FALSE
)

write.csv(
  mixed_paths,
  file.path(out_dir, "parallel_sequential_paths.csv"),
  row.names = FALSE
)

write.csv(
  mixed_intermediate_paths,
  file.path(out_dir, "parallel_sequential_intermediate_paths.csv"),
  row.names = FALSE
)

saveRDS(
  list(
    parallel_fit = parallel_fit,
    sequential_fit = sequential_fit,
    mixed_fit = mixed_fit,
    mixed_intermediate_fit = mixed_intermediate_fit
  ),
  file.path(out_dir, "asd_sequential_zentangler_fits.rds")
)
```

# Interpreting The Output

For sequential and mixed route sets, the main path table includes:

| Column | Meaning |
|---|---|
| `route_set_type` | `sequential` or `parallel_sequential` |
| `path_template` | user-specified route name |
| `view_route` | modality-level route |
| `route_type` | `sequential` or `parallel_single_view` |
| `mediator_path` | feature-level mediator path |
| `a` | exposure-to-first-mediator coefficient |
| `d_product` | product of mediator-to-mediator link coefficients |
| `b` | terminal mediator-to-outcome coefficient |
| `sequential_score` | `a * d_product * b` |
| `fusion_mode` | terminal fusion mode: `early`, `intermediate`, or `late` |
| `p_primary` | conservative path-level screening p-value |
| `q_primary` | global FDR-adjusted path-level value |

For the pure parallel model, use the usual Zentangler helpers:

```{r}
head(zentangler_all_mediators(parallel_fit))
zentangler_threshold_summary(parallel_fit, q_threshold = seq(0.05, 0.25, by = 0.05))
```
