---
title: "Zentangler: multiview parallel mediation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zentangler: multiview parallel mediation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

Zentangler is an MAE-first R package for generic multiview parallel mediation
analysis. The package is designed for studies where an exposure variable `X` may
influence many candidate mediators distributed across multiple omics or assay
views, and those mediators may in turn be associated with an outcome `Y`.

The active method focuses on the parallel multiview mediation setting:

```text
X -> mediator in view 1 -> Y
X -> mediator in view 2 -> Y
...
X -> mediator in view K -> Y
```

The package is intentionally generic. A view can be microbiome, metabolome,
proteome, transcriptome, pathway abundance, clinical biomarkers, imaging-derived
features, or any other numeric feature matrix. Zentangler does not assume exactly
two views.

The public user-facing input is a `MultiAssayExperiment` object. This is useful
for Bioconductor package development because it keeps assay data, sample mapping,
and phenotype variables in one coordinated container.

# Statistical target

For each mediator feature `j` in view `k`, Zentangler estimates two main pieces
of evidence.

The first piece is the A-path:

```text
Does X associate with mediator j in view k?
```

This gives an A-path coefficient `a_hat` and A-path p-value `p_a`.

The second piece is the B-path:

```text
After accounting for X and other selected mediators, does mediator j associate
with Y?
```

This gives a B-path coefficient `b` and B-path p-value `p_b`.

The main mediation score is:

```text
score_j = a_hat_j * b_j
```

A positive score means the A-path and B-path have the same sign. A negative score
means the A-path and B-path have opposite signs. Larger absolute scores are used
as a ranking measure.

Important scientific distinction: `score = a*b` remains the mediator-level
screening and ranking score. Zentangler separately reports causal direct,
indirect, and total effect summaries using an active-set parametric
g-computation layer. These causal summaries are conditional on the selected
mediator set and on the usual mediation identification assumptions; they are not
a replacement for sensitivity analysis or a final selective-inference proof.

# Input data contract

The primary input is a `MultiAssayExperiment`.

Zentangler expects:

1. One or more experiments in `experiments(mae)`.
2. Each experiment represents one view.
3. Features are numeric.
4. Samples can be aligned to `colData(mae)`.
5. `colData(mae)` contains the exposure variable `x_var`.
6. `colData(mae)` contains the outcome variable `y_var`.
7. Optional covariates are also stored in `colData(mae)`.

Most Bioconductor assays are stored as features x samples. Zentangler extracts
these into sample x feature matrices internally. Plain matrix-like experiments
are also supported when sample IDs can be mapped.

# Creating a small MAE

The example below creates a small toy `MultiAssayExperiment`. In real analyses,
you would replace these simulated matrices with your own assay data.

```{r}
library(Zentangler)
library(MultiAssayExperiment)
library(SummarizedExperiment)

set.seed(1)

n <- 80
sample_ids <- paste0("sample_", seq_len(n))

pheno <- DataFrame(
  X = rbinom(n, 1, 0.5),
  Y = rnorm(n),
  age = rnorm(n, mean = 50, sd = 10),
  subject_id = sample(paste0("subject_", 1:40), n, replace = TRUE),
  row.names = sample_ids
)

view1 <- matrix(rnorm(40 * n), nrow = 40, ncol = n)
view2 <- matrix(rnorm(60 * n), nrow = 60, ncol = n)

rownames(view1) <- paste0("view1_feature_", seq_len(nrow(view1)))
rownames(view2) <- paste0("view2_feature_", seq_len(nrow(view2)))
colnames(view1) <- sample_ids
colnames(view2) <- sample_ids

mae <- MultiAssayExperiment(
  experiments = ExperimentList(
    view1 = SummarizedExperiment(assays = list(abundance = view1)),
    view2 = SummarizedExperiment(assays = list(abundance = view2))
  ),
  colData = pheno
)
```

# Running the main model

A minimal fit uses a method preset.

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  method_preset = "fast_lasso",
  sis_n = 25,
  fdr_method = "BH",
  fdr_scope = "global"
)
```

The `method_preset` argument is meant to make the API safer and more stable. It
lets users select a coherent method configuration without manually setting every
low-level option.

You can inspect available presets with:

```{r}
zentangler_method_presets()
```

Current presets include:

```text
custom
fast_lasso
elastic_net
longitudinal_maaslin2
full_exploratory
```

`custom` keeps user-specified arguments. The other presets set common defaults
for A-stage modeling, screening, fusion, penalty choice, and B-stage inference.

# Stage 1: A-path estimation

The A-stage asks whether each candidate mediator is associated with the exposure.
This is done separately within each view.

For mediator `j` in view `k`, the basic model is:

```text
M_jk = alpha_jk + a_jk * X + covariate terms + error_jk
```

The coefficient `a_jk` is the A-path estimate. The p-value testing whether
`a_jk = 0` is stored as `p_a`.

Zentangler currently provides two A-stage options.

## HIMA-like linear A-stage

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  a_stage_model = "lm",
  sis_n = 25,
  fusion_mode = "early"
)
```

This option uses transparent univariate linear modeling. It is simple, fast, and
close in spirit to the first screening step used in HIMA-style pipelines.

## MaAsLin2 A-stage

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  a_stage_model = "maaslin2",
  maaslin2_random_effect = "subject_id",
  maaslin2_normalization = "NONE",
  maaslin2_transform = "NONE",
  maaslin2_analysis_method = "LM",
  sis_n = 25,
  fusion_mode = "early"
)
```

This option is useful when the A-stage needs MaAsLin2-style modeling, especially
when repeated measures or subject-level random effects are important. In this
mode, covariates and random effects are modeled directly by MaAsLin2 rather than
through the internal residualization path.

# Stage 2: Screening

High-dimensional mediation can involve hundreds or thousands of mediators per
view. Zentangler therefore supports screening before fitting the Y model.

The default screening strategy is SIS-style filtering:

```text
Keep the top N mediators per view.
```

The ranking can be based on either the magnitude of the A-path estimate or the
A-path p-value.

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  screen_method = "sis",
  sis_n = 50,
  sis_rank = "abs_a"
)
```

To turn this off and pass all filtered numeric mediators into the B-stage:

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  screen_method = "none"
)
```

Turning off screening can increase power when sample size is large enough, but it
also makes the Y-stage larger and can be less stable in small sample settings.

# Stage 3: Multiview B-stage fusion

After A-stage estimation and optional screening, Zentangler fits a Y-stage model
that includes `X`, selected mediators from the views, and optional covariates.

The conceptual model is:

```text
Y = direct X term + selected mediator terms + covariate terms + error
```

The mediator coefficient from this stage is the B-path estimate `b`.

Zentangler currently supports three fusion modes.

## Early fusion

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fusion_mode = "early",
  glmnet_alpha = 1,
  lambda_choice = "lambda.1se"
)
```

Early fusion concatenates all selected mediators from all views into one design
matrix and fits one penalized model. This lets all views compete directly inside
a single model. It is simple and often a good first option.

`glmnet_alpha = 1` gives lasso. Values between 0 and 1 give elastic net. For
example:

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fusion_mode = "early",
  glmnet_alpha = 0.5
)
```

## Intermediate fusion

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fusion_mode = "intermediate",
  coop_rho = 0.2
)
```

Intermediate fusion uses a cooperative learning style update. Instead of simply
concatenating all mediators, it encourages agreement across view-specific fitted
signals. This is useful when the goal is to borrow strength across views while
still keeping view-specific mediator coefficients.

At present, the intermediate mode uses a cooperative lasso-style Gaussian Y-stage
implementation. The `glmnet_alpha` elastic-net setting is currently applied to
early and late fusion, not to the intermediate cooperative update.

## Late fusion

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fusion_mode = "late",
  glmnet_alpha = 1
)
```

Late fusion fits view-specific penalized models first and then combines the
view-level fitted signals in a meta model. This can help preserve view-level
structure and can be useful when views have very different feature scales or
feature counts.

# Stage 4: B-path inference

The default B-stage inference is HIMA-style de-biased lasso inference:

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  b_inference = "debiased_lasso"
)
```

The purpose is to obtain approximate p-values for mediator B-path coefficients
after sparse model fitting. The package also includes a refit option:

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  b_inference = "refit"
)
```

The refit option uses the selected active set and computes ordinary refit-based
p-values. This is simpler to explain, but can be less stable if the selected
active set is too large relative to sample size.

# Stage 5: Combining mediator evidence

For each mediator, Zentangler reports:

```text
a_hat      A-path estimate
p_a        A-path p-value
b          B-path estimate
p_b        B-path p-value
score      a_hat * b
abs_score  abs(score)
p_primary  primary mediator evidence p-value
q_primary  selected final FDR-adjusted p_primary
```

The main ranking score is `score = a_hat * b`.

The current p-value combination is conservative. In the default output,
`p_primary` follows the joint A/B evidence field, which is currently based on the
larger of the A-path and B-path p-values when both are available. This asks for
both sides of the mediation pattern to show evidence.

This is useful for screening and ranking. It should not be described as final
post-selection-valid causal inference.

# Multiple-testing scope

Zentangler separates two choices that are easy to confuse:

1. Which FDR correction method should be used?
2. What family of hypotheses should the final q-values be adjusted over?

The correction method is controlled by `fdr_method`:

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fdr_method = "BH"
)
```

`fdr_method = "BH"` uses Benjamini-Hochberg. This is the usual default.
`fdr_method = "BY"` uses Benjamini-Yekutieli, which is more conservative and can
be useful when dependence across tests is a major concern.

The final q-value family is controlled by `fdr_scope`:

```{r}
fit_global <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fdr_scope = "global"
)

fit_within <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fdr_scope = "within_view"
)
```

`fdr_scope = "global"` adjusts the final primary p-values across all mediators
from all assay views. This treats the full multiview mediator search as one
multiple-testing family.

`fdr_scope = "within_view"` adjusts the final primary p-values separately inside
each assay view. This treats each view as its own multiple-testing family.

The combined mediator table keeps both q-value columns regardless of which scope
is selected:

```text
q_primary_global       final q-value adjusted across all views
q_primary_within_view  final q-value adjusted separately within each view
q_primary              q-value used by the current fit, based on fdr_scope
```

This makes it possible to compare global and within-view calls after fitting,
without changing the A-stage or B-stage model.

# Optional bootstrap uncertainty

Zentangler can run bootstrap resampling to add uncertainty summaries for mediator
scores and effect summaries.

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fusion_mode = "early",
  bootstrap_repeats = 100,
  bootstrap_ci_level = 0.95,
  seed = 1
)
```

For repeated-measure or longitudinal data, use `bootstrap_id` to resample at the
subject level rather than treating repeated observations as independent samples.

```{r}
fit <- fit_multiview_parallel_zentangler(
  mae = mae,
  x_var = "X",
  y_var = "Y",
  fusion_mode = "early",
  bootstrap_repeats = 100,
  bootstrap_id = "subject_id"
)
```

Subject-level bootstrap is usually more appropriate when multiple samples come
from the same subject.

# Inspecting outputs

The full mediator table is available as:

```{r}
all_mediators <- zentangler_all_mediators(fit)
head(all_mediators)
```

The top-ranked mediators are available with:

```{r}
zentangler_top_mediators(fit, n = 20)
```

Active mediators can be extracted by BH q-value threshold:

```{r}
zentangler_active_mediators(fit, q_threshold = 0.25)
```

View-level summaries are available with:

```{r}
zentangler_view_summary(fit, q_threshold = 0.25)
```

A model-level summary is available with:

```{r}
zentangler_model_summary(fit, q_threshold = 0.25)
```

A compact summary bundle is available with:

```{r}
summary <- summarize_zentangler(fit, q_threshold = 0.25)
names(summary)
```

The summary helper can also evaluate several q-value thresholds from the same
fitted model:

```{r}
summary <- summarize_zentangler(
  fit,
  q_threshold = seq(0.05, 0.25, by = 0.05)
)

summary$threshold_summary
```

This does not refit the model. It only re-counts active mediators under each
threshold and each available q-value column.

The fit object also contains a `diagnostics` field:

```{r}
fit$diagnostics
```

This includes sample counts, view counts, feature counts before and after
filtering, screened mediator counts, active B-path counts, runtime, R version,
and bootstrap failures if any occurred.

# Effect decomposition

The fit object includes an `effect_decomposition` table:

```{r}
fit$effect_decomposition
```

This reports selected-active-set natural-direct-effect-style, natural-indirect-effect-style,
and total-effect-style summaries. For binary exposures, the default contrast is
the lower observed exposure level versus the higher observed exposure level. For
continuous exposures, the default contrast is the 25th versus 75th percentile,
unless `effect_x0` and `effect_x1` are supplied by the user.

For Gaussian outcomes, effects are reported on the mean-difference scale. For
binary outcomes, effects are reported on the risk-difference scale by averaging
predicted probabilities under the relevant counterfactual mediator values.

# Running simulations

Zentangler includes simulation wrappers for InterSIM/SIMMBA-style MAE simulation
studies.

```{r}
res <- run_intersim_zentangler(
  nrep = 10,
  nsample = 600,
  fusion_modes = c("early", "intermediate", "late"),
  screen_method = "sis",
  sis_n = 200,
  fdr_method = "BH",
  fdr_scope = "both",
  q_threshold = seq(0.05, 0.25, by = 0.05),
  lambda_choice = "lambda.min",
  glmnet_alpha = 1
)

res$summary
```

The simulation wrapper returns:

```text
summary    average performance across simulation replicates
detail     replicate-level performance
fits       fitted Zentangler objects
truth      simulation truth tables
errors     model failures, if any
settings   simulation and model settings
```

When `q_threshold` is a vector, Zentangler fits each simulated dataset once and
then evaluates active mediator recovery at each threshold. When
`fdr_scope = "both"`, the simulation summary includes both global and within-view
q-value families. This is useful for HPC runs because it avoids launching
separate jobs that differ only by the active-mediator q cutoff.

Common simulation metrics include:

```text
n_active          number of selected active mediators
true_active       number of selected mediators that are truly active
false_active      number of selected mediators that are false positives
precision         true_active / n_active
recall            true_active / total true mediators
fdr               false_active / n_active
top50_true        number of true mediators in the top 50 ranked mediators
top50_precision   top50_true / 50
```
