---
title: "Zentangler_ISMB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zentangler_ISMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

This short tutorial uses an ASD example for comparing:

1. `HIMA2`, run one omics view at a time.
2. `Zentangler`, run across all ASD omics views together.

This tutorial uses:

- the ASD example data already shipped in `data/`
- one symptom outcome: `abc`
- one time contrast: `T0` versus `T1`
- a continuous outcome only


# Data Used

The modalities used are:

```text
Species
KOs
Fecal Metabolites
Plasma Metabolites
```

The exposure is treatment time:

```text
X = 0 for baseline `T0`
X = 1 for T1
```

The outcome is the `abc` symptom score.

# Packages

```{r}
library(Zentangler)
library(MultiAssayExperiment)
library(SummarizedExperiment)
library(S4Vectors)

if (!requireNamespace("HIMA", quietly = TRUE)) {
  stop("Install HIMA before running the HIMA2 part of this tutorial.")
}
```



# Load The RData


```{r}
load('/Users/nalinarora/Desktop/Zentangler/data/asd_ABC_t1_baseline.RData')

mae_t1_abc

```

# Run Zentangler

For a first pass, `late` fusion is a practical choice because it lets each omics
view keep its own signal structure before combining the view-level predictions.

```{r}
zt_fit <- fit_multiview_parallel_zentangler(
  mae_t1_abc,
  y_var = "Y",
  study_design = "time",
  time_var = "timepoint",
  time_ref = "T0",
  time_compare = "T1",
  covariates = NULL,
  screen_method = "sis",
  sis_n = 50,
  sis_rank = "abs_a",
  fusion_mode = "early",
  glmnet_alpha = 0.5,
  y_family = "gaussian",
  fdr_method = "BH",
  fdr_scope = "global",
  fill_nonfinite_zero = TRUE,
  seed = 1
)

zentangler_model_summary(zt_fit)
zentangler_view_summary(zt_fit, q_threshold = 0.25)
zentangler_top_mediators(zt_fit, n = 15)
zentangler_active_mediators(zt_fit, q_threshold = 0.25)
```

# Run HIMA2 One View At A Time

HIMA2 is a single-view method, so we loop over the ASD modalities one by one.

```{r}
mae_block <- function(mae, view) {
  extract_block <- getFromNamespace("zentangler_mae_extract_block", "Zentangler")
  extract_block(
    mae,
    experiment_name = view,
    fill_nonfinite_zero = TRUE
  )
}

run_hima2_one_view <- function(M, pheno, penalty = "DBlasso", outcome_family = "gaussian") {
  ns <- asNamespace("HIMA")
  fun_name <- if (exists("hima", envir = ns, inherits = FALSE)) "hima" else "hima2"
  fun <- get(fun_name, envir = ns)
  argn <- names(formals(fun))
  data_M <- as.data.frame(M, check.names = FALSE)

  if ("formula" %in% argn) {
    call_args <- list(formula = Y ~ X)
    if ("data.pheno" %in% argn) call_args$data.pheno <- pheno
    if ("data.M" %in% argn) call_args$data.M <- data_M
    if ("mediator.type" %in% argn) call_args$mediator.type <- "gaussian"
    if ("mediator.family" %in% argn) call_args$mediator.family <- "gaussian"
    if ("outcome.family" %in% argn) call_args$outcome.family <- "gaussian"
    if ("penalty" %in% argn) call_args$penalty <- penalty
    if ("scale" %in% argn) call_args$scale <- FALSE
    if ("verbose" %in% argn) call_args$verbose <- FALSE
    if ("topN" %in% argn) call_args$topN <- ncol(data_M)
    if ("sigcut" %in% argn) call_args$sigcut <- 1
    if ("Sigcut" %in% argn) call_args$Sigcut <- 1
    return(do.call(fun, call_args))
  }

  call_args <- list(X = pheno$X, Y = pheno$Y, M = as.matrix(M))
  if ("penalty" %in% argn) call_args$penalty <- penalty
  if ("topN" %in% argn) call_args$topN <- ncol(M)
  if ("sigcut" %in% argn) call_args$sigcut <- 1
  if ("Sigcut" %in% argn) call_args$Sigcut <- 1
  do.call(fun, call_args)
}

standardize_hima2_table <- function(fit, view) {
  tab <- tryCatch({
    if (inherits(fit, "hima") && is.list(fit)) {
      lens <- vapply(fit, length, integer(1))
      keep <- names(fit)[lens == max(lens) & lens > 0]
      as.data.frame(fit[keep], check.names = FALSE, stringsAsFactors = FALSE)
    } else {
      as.data.frame(fit, check.names = FALSE)
    }
  }, error = function(e) data.frame())
  if (nrow(tab) == 0) return(tab)

  if ("ID" %in% names(tab)) {
    tab$mediator <- as.character(tab$ID)
  } else {
    tab$mediator <- rownames(tab)
  }
  rownames(tab) <- NULL
  names(tab) <- make.names(names(tab), unique = TRUE)

  p_col <- intersect(c("p.value", "p.value.", "p", "P.value", "raw.p", "raw_p"), names(tab))
  q_col <- intersect(c("BH.FDR", "BH_FDR", "q.value", "qvalue", "FDR"), names(tab))
  a_col <- intersect(c("alpha", "a", "alpha_hat"), names(tab))
  b_col <- intersect(c("beta", "b", "beta_hat"), names(tab))
  ab_col <- intersect(c("alpha.beta", "alpha.beta.", "alpha.beta.."), names(tab))

  tab$p_primary <- if (length(p_col) > 0) as.numeric(tab[[p_col[[1]]]]) else NA_real_
  tab$q_hima_reported <- if (length(q_col) > 0) as.numeric(tab[[q_col[[1]]]]) else NA_real_
  tab$a <- if (length(a_col) > 0) as.numeric(tab[[a_col[[1]]]]) else NA_real_
  tab$b <- if (length(b_col) > 0) as.numeric(tab[[b_col[[1]]]]) else NA_real_
  tab$score <- if (length(ab_col) > 0) as.numeric(tab[[ab_col[[1]]]]) else tab$a * tab$b
  tab$view <- view
  tab
}

run_hima2_mae <- function(mae, x_var = "X", y_var = "Y") {
  pheno <- as.data.frame(SummarizedExperiment::colData(mae))
  pheno$X <- suppressWarnings(as.numeric(as.character(pheno[[x_var]])))
  pheno$Y <- suppressWarnings(as.numeric(as.character(pheno[[y_var]])))
  pheno <- pheno[is.finite(pheno$X) & is.finite(pheno$Y), , drop = FALSE]

  results <- list()

  for (view in names(MultiAssayExperiment::experiments(mae))) {
    M <- mae_block(mae, view)
    common <- intersect(rownames(pheno), rownames(M))
    M <- M[common, , drop = FALSE]
    pheno_view <- pheno[common, , drop = FALSE]

    if (length(unique(pheno_view$X)) < 2 || nrow(pheno_view) < 8 || ncol(M) < 2) {
      next
    }

    fit <- run_hima2_one_view(M, pheno_view)
    results[[view]] <- standardize_hima2_table(fit, view)
  }

  out <- if (length(results) > 0) do.call(rbind, results) else data.frame()
  if (nrow(out) > 0) {
    out$q_primary_global <- p.adjust(out$p_primary, method = "BH")
  }
  rownames(out) <- NULL
  out
}

hima2_tab <- run_hima2_mae(mae_t1_abc)
head(hima2_tab[order(hima2_tab$q_primary_global), ], 15)
```

# Simple HIMA2 Versus Zentangler Comparison

Use the same ASD slice and compare the top ranked mediators from both methods.

```{r}
zt_tab <- zentangler_all_mediators(zt_fit)
zt_top <- zt_tab[order(zt_tab$q_primary), c("omics", "mediator", "score", "q_primary_global")]
hima_top <- hima2_tab[order(hima2_tab$q_primary_global), c("view", "mediator", "p_primary", "q_primary_global")]

head(zt_top, 10)
head(hima_top, 10)
```

Interpretation:

- `HIMA2` asks whether each modality contains mediators when analyzed alone.
- `Zentangler` asks whether mediators remain important after joint multiview
  screening and fusion.
- If a feature ranks highly in both methods, it is often a useful follow-up
  candidate.

# Compare Fusion Strategies

| Fusion mode | Idea | Practical note |
| --- | --- | --- |
| `early` | stack selected mediators from all views into one model | easiest to explain |
| `intermediate` | encourage view-specific models to agree | Gaussian outcomes only |
| `late` | fit each view first, then combine view-level predictions | helpful when views differ a lot |

```{r}
fusion_modes <- c("early", "intermediate", "late")

fusion_fits <- lapply(fusion_modes, function(mode) {
  fit_multiview_parallel_zentangler(
    mae_t1_abc,
    y_var = "Y",
    study_design = "time",
    time_var = "timepoint",
    time_ref = "T0",
    time_compare = "T1",
    screen_method = "sis",
    sis_n = 50,
    sis_rank = "abs_a",
    fusion_mode = mode,
    glmnet_alpha = 0.5,
    y_family = "gaussian",
    fdr_method = "BH",
    fdr_scope = "global",
    fill_nonfinite_zero = TRUE,
    seed = 1
  )
})

names(fusion_fits) <- fusion_modes

fusion_summary <- do.call(rbind, lapply(names(fusion_fits), function(mode) {
  top_tab <- zentangler_top_mediators(fusion_fits[[mode]], n = 5)
  data.frame(
    fusion_mode = mode,
    n_active_q0p25 = nrow(zentangler_active_mediators(fusion_fits[[mode]], q_threshold = 0.25)),
    top_mediators = paste(top_tab$mediator, collapse = "; "),
    stringsAsFactors = FALSE
  )
}))

fusion_summary
```

# Top 5 DAGs

The code below keeps mediators with `q_primary < 0.25`, then plots the top 5
ranked by `score`.

```{r, fig.width=14, fig.height=16, out.width='100%', fig.align='center'}
zt_sig <- subset(
  zentangler_all_mediators(zt_fit),
  is.finite(q_primary) & q_primary < 0.25
)

zt_sig <- zt_sig[order(zt_sig$score, decreasing = TRUE, na.last = TRUE), , drop = FALSE]
zt_top5 <- head(zt_sig, 5)

wrap_node_text <- function(x, width = 22) {
  paste(strwrap(x, width = width), collapse = "\n")
}

format_effect <- function(x) {
  ifelse(is.finite(x), formatC(x, digits = 3, format = "f"), "NA")
}

draw_box <- function(xleft, ybottom, xright, ytop, label,
                     fill = "#1155cc", border = "#8fb3ff", cex = 1.1) {
  rect(xleft, ybottom, xright, ytop, col = fill, border = border, lwd = 2)
  text((xleft + xright) / 2, (ybottom + ytop) / 2, label,
       col = "white", cex = cex, font = 2)
}

draw_labeled_arrow <- function(x0, y0, x1, y1, label, label_x, label_y,
                               col = "#1155cc", lwd = 2.5, cex = 1.2) {
  arrows(x0, y0, x1, y1, length = 0.1, lwd = lwd, col = col)
  text(label_x, label_y, labels = label, cex = cex, font = 2, col = "black")
}

plot_mediator_dag <- function(row_i, rank_i, outcome_label = "ABC") {
  mediator_label <- paste0(
    wrap_node_text(as.character(row_i$mediator), width = 24),
    "\n(",
    gsub("_", " ", as.character(row_i$omics), fixed = TRUE),
    ")"
  )

  plot.new()
  plot.window(xlim = c(0, 1), ylim = c(0, 1))
  title(main = paste0(rank_i, "."), adj = 0.04, cex.main = 1.8, font.main = 2)

  draw_box(0.03, 0.08, 0.36, 0.26, "10 weeks vs. baseline", cex = 1.25)
  draw_box(0.27, 0.62, 0.69, 0.90, mediator_label, cex = 1.05)
  draw_box(0.70, 0.10, 0.96, 0.28, outcome_label, cex = 1.25)

  draw_labeled_arrow(
    0.17, 0.26, 0.40, 0.62,
    format_effect(as.numeric(row_i$a)),
    label_x = 0.24, label_y = 0.45
  )
  draw_labeled_arrow(
    0.58, 0.62, 0.83, 0.28,
    format_effect(as.numeric(row_i$b)),
    label_x = 0.76, label_y = 0.44
  )
  draw_labeled_arrow(
    0.36, 0.17, 0.70, 0.17,
    format_effect(as.numeric(row_i$score)),
    label_x = 0.53, label_y = 0.24
  )
}

old_par <- par(no.readonly = TRUE)
on.exit(par(old_par), add = TRUE)

par(
  mfrow = c(3, 2),
  mar = c(1.2, 1.2, 2.2, 1.2),
  oma = c(0, 0, 1, 0),
  xaxs = "i",
  yaxs = "i"
)

for (i in seq_len(nrow(zt_top5))) {
  plot_mediator_dag(
    row_i = zt_top5[i, , drop = FALSE],
    rank_i = i,
    outcome_label = "ABC"
  )
}

if (nrow(zt_top5) < 6) {
  plot.new()
}
```

```{r}
sessionInfo()
```